PoVaTiM: A Possibilistic Valid-Time Model \\ \vspace*{0.04truein}

### 2012-06-13 ijcis-pos-model/sources/timeRepresentation.tex
version sent to Olga and Nicolas

This section is devoted to specify the representation of time within the framework of the possibility theory. First of all, the specification for a single ill-known temporal point shall be explained. Then, the formal specification and the related constraints are given for an ill-known valid-time interval.

\subsection{\label{subsec:ill-known-point-rep}Ill-known time point}
An ill-known time point $X$ is a precise time point that, for some reason, it is not fully specified. Note that $X$ has only one possible value but that value is unspecified.

\begin{definition}
\label{def:ill-known-time-point}
\textbf{Ill-known time point.}\\
Consider a time domain $\T$, uncertainty about the values of the ill-known time point $X$ is given by the possibility distribution $\pi_X$:
\begin{equation}
\label{eq:ill-known-time-point}
\Pi(X) = \pi_X(t) \in \left[0,1\right] , t \in \T
\end{equation}
\end{definition}

It is also possible to specify an ill-known time point by a convex combination of ill-known constraints, (Definition \ref{def:convex-combination-ill-known-constraints}) as shown by equation \eqref{eq:ill-known-value-def-by-const-unc}.

@@

\subsubsection{\label{subsubsec:ill-known-time-datatypes}Datatypes}
The data type for the representation of an ill-known time point allows the representation of the values shown in Table \ref{tbl:time-point-types}
%The datatypes for an  $X$ are shown in Table \ref{}.
\vglue13pt
%\begin{table}[htbp]
\tcap{\label{tbl:time-point-types}Values for the time point data type.}
\centerline{\small TIME POINT DATA TYPE}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c p{0.4\columnwidth} p{0.3\columnwidth}}\\
Subtype & Value & Representation \\
\hline
1 & A single time point &  $1/x, x \in \T$\\
2 & A possibility distribution in the numeric domain & A fuzzy number or a fuzzy interval.\\
3 & An unknown value & \textbf{UNKNOWN}$= \lbrace1/t,\forall t \in \T \rbrace$\\
4 & An undefined value & \textbf{UNDEFINED}$= \lbrace 0/t,\forall t \in \T \rbrace$\\
5 & A null value & \textbf{NULL} $=\lbrace 1/$Unknown, 1/Undefined $\rbrace$\\
\hline\\
\end{tabular}
}

%Example of different uses of the ill-known time value.
\begin{example}
Consider a historical database with data from diplomatic documents. The following fields are stored: the digital identifier \emph{ID} which is the primary key  and the estimated time when the document was sent (field \emph{Date}). Table \ref{tbl:sample-time-point} contains some example data from this database. 
\end{example}

@@

\begin{definition}
\textbf{Ill-known time interval}\\
An ill-known time interval denoted by $\left[X, Y\right]$ is a precise time interval whose boundaries are not precisely known. The ill-known interval is defined by two ill-known points, namely $X$,$Y$.
In order to know if all the points in the crisp time interval $I=\left[a, b\right]$ are within the boundaries of the ill-known time interval $\left[X, Y\right]$ we define the following two ill-known constraints $C_{1}, C_{2}$ 
\begin{eqnarray}
\label{eq:constraint-c1}
C_1\stackrel{\triangle}{=}\left(\geq,X\right)  \\
\label{eq:constraint-c2}
C_2\stackrel{\triangle}{=}\left(\leq,Y\right)
\end{eqnarray}
Note that the boolean operator applied to the constraints is $\bool=\wedge$. Hence, the construct of both constraints is $C_1 \wedge C_2$
This means that we want to know if all the points in the interval $I$ are greater or equal to $X$ and, at the same time, smaller or equal to $Y$.
\end{definition}

%% equations for the poss and nec!!
The simplified equations for the possibility and the necessity measures are~\cite{Pons2011}:
\begin{eqnarray}
\label{eq:interval-pos}
\Pos\left(\lambda([a,b])\right)&=&\\
\nonumber
\min\bigg(\sup_{a\leq w}\pi_{X}(w),\sup_{b\geq w}\pi_{Y}(w)\bigg)\\
\label{eq:interval-nec}
\Nec\left(\lambda([a,b])\right)&=&\\
\nonumber
\min\bigg(\inf_{a>w}1-\pi_{X}(w),\inf_{b<w}1-\pi_{Y}(w)\bigg).
\end{eqnarray}

@@

\subsubsection{\label{subsubsec:representation-semi-open}Representation of semi-open time intervals}
As mentioned before, the problem resides in the representation of these kind of interval. In order to get a short representation for these intervals, two constants are defined: $FB$ \emph{From the Beginning}: it is only valid for the left value of the interval and $UC$ \emph{Until Changed} (this name is usual in the temporal database community, see ~\cite{Jensen1994}) which is only valid for the right value of the interval. Note that these two constants are aliases for a function that obtains both possibility and necessity measures.

Because of the ill-known constraints $C_{1}, C_{2}$, a function called \emph{Open} should be defined in order to deal with a proper representation of these intervals:

\begin{definition}
\label{def:open-func}
$Open(C)$\\
The function $\Open(C)$ provides both possibility and necessity measures for all the points in the open part of a semi-open ill-known time interval.
The parameter $C$ is an ill-known constraint $C\stackrel{\triangle}{=}\left(R,T\right)$ as defined in \ref{subsec:interval-evaluation-by-ill-known-constraints}, in the equation \eqref{eq:ill-known-constraint} and in~\cite{Pons2011}.
The possibility and necessity measures are defined by:
\begin{eqnarray}
\label{eq:open-pos-nec}
\Pos\left(\Open(C)\right) &=& \Big(\sup_{r\in\T \Rp\  w}\pi_{T}(w)\Big)\\
\Nec\left(\Open(C)\right) &=& \Big(\inf_{r\in\T \Rn\  w}1-\pi_{T}(w)\Big)
\end{eqnarray}
Where the values for the binary relations $\Rp$ and $\Rn$ are shown in Table \ref{tbl:open-pos-nec-rels}.
\end{definition}

\vglue13pt
%\begin{table}[htbp]
\tcap{Relations for the $\Open(C)$ function. Depending on the relation $R = \left(\leq, <, >, \geq\right)$ in the constraint $C$, the values for $\Rp$ and $\Rn$ are shown.}
\centerline{\small Relations}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c c c}\\
\hline
Constraint & $\Rp$ & $\Rn$\\ \hline
$C\stackrel{\triangle}{=}\left(<,T\right)$ & $>$ & $\leq$ \\
$C\stackrel{\triangle}{=}\left(\leq,T\right)$ & $\geq$ & $<$ \\
$C\stackrel{\triangle}{=}\left(>,T\right)$ & $<$ & $\geq$ \\
$C\stackrel{\triangle}{=}\left(\geq,T\right)$ & $\leq$ & $>$ \\
\hline\\
\end{tabular}
\label{tbl:open-pos-nec-rels}
}

@@

\begin{example}
\label{ex:open-time-interval}
Consider an ill-known time interval given by $\left[\FB, Y \right]$. Consider also that in this case, $Y= \left[15/10/2012,3,4 \right] $. Figure \ref{fig:example-open-time-interval} shows the representation for $Y$. The user wants to obtain the possibility and the necessity measures for the $\FB$ part of the interval are:
\end{example}

\begin{eqnarray}
\nonumber
\FB = \Open(C_2) \mbox{ with } C_2\stackrel{\triangle}{=}\left(\leq,Y\right)\\
\nonumber
\Pos\left(\Open(C)\right) = \Big(\sup_{r\in\T \geq\  w}\pi_{Y}(w)\Big)\\
\nonumber
\Nec\left(\Open(C)\right) = \Big(\inf_{r\in\T <\  w}1-\pi_{Y}(w)\Big)
\end{eqnarray}

@@

\subsubsection{\label{subsubsec:ill-known-time-interval-datatypes}Datatypes}
In order to represent properly an ill-known time interval in a database, some datatypes are needed. Because of the ill-known constraints, not all the combinations of datatypes for each ill-known time point (see Table \ref{tbl:time-point-types}) are allowed. Table \ref{tbl:time-interval-types} shows all the possible combinations for the datatypes for an ill-known time interval denoted by $\left[X, Y\right]$.

@@

\subsection{\label{subsec:fuzzy-allen-relations}Allen's relations}
The Allen's relations can be represented by a Boolean combination of ill-known constraints. The constructs for the constraints are shown in Table \ref{tbl:fuzzy-allen-relations}. The possibility and necessity measures are obtained for each Boolean function as explained in Section \ref{subsec:interval-evaluation-by-ill-known-constraints} in equations \eqref{eq:conjunctive1}-\eqref{eq:negation}.

\vglue13pt
%\begin{table}[htbp]
\tcap{\label{tbl:fuzzy-allen-relations}Constructs of constraints related to their respective Allen relations, as used in the presented work. In this table, the ill-known time interval $J = \left[X, Y\right]$ in a record $r$ has a start point described by possibilistic variable $X$ and an end point described by possibilistic variable $Y$. The crisp time interval in the user's temporal demand is denoted $I$.}
\centerline{\small Relations}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{p{0.2\columnwidth} p{0.8\columnwidth} }\\
\hline
Allen Relation & Construct of constraints\\
\hline
I before J & $C_1 \triangleq \left(<,X\right)$ \\
\hline
I equal J & $\left(C_1 \triangleq \left(\geq,X\right)\right) \wedge \neg \left(C_2 \triangleq \left(\neq,X\right) \right) \wedge \left(C_3 \triangleq \left(\leq,Y\right) \right) \wedge \neg \left(C_4 \triangleq \left(\neq,Y\right)\right)$ \\
\hline
I meets J & $\left(C_1 \triangleq \left(\leq,X\right)\right) \wedge \neg \left(C_2 \triangleq \left(\neq,X\right) \right)$ \\
\hline
I overlaps J & $\left(C_1 \triangleq \left(<,Y\right)\right) \wedge \neg \left(C_2 \triangleq \left(\leq,X\right) \right) \wedge \neg \left(C_3 \triangleq \left(\geq,X\right) \right)$ \\
\hline
I during J & $\left( \left (C_1 \triangleq \left(>,X\right) \right) \wedge \left(C_2 \triangleq \left(\leq,Y\right) \right) \right) \vee \left( \left(C_3 \triangleq \left(\geq,X\right) \right) \wedge  \left(C_4 \triangleq \left(<,Y\right) \right) \right)$ \\
\hline
I starts J & $\left(C_1 \triangleq \left(\geq,X\right) \right) \wedge \neg \left(C_2\triangleq \left(\neq,X\right)\right)$ \\
\hline
I finishes J & $\left(C_1 \triangleq \left(\leq,Y\right) \right) \wedge \neg \left(C_2\triangleq \left(\neq,Y\right)\right)$ \\
\hline\\
\end{tabular}
}

>>>>

This section is devoted to specify the representation of time within the framework of the possibility theory. First of all, the specification for a single ill-known temporal point will be explained. Then, the formal specification and the related constraints are given for an ill-known valid-time interval.

\subsection{\label{subsec:ill-known-point-rep}Ill-known time point}
An ill-known time point $X$ is a precise time point that, for some reason, is not fully specified. Note that $X$ has only one possible value but that value is unspecified.

\begin{definition}
\label{def:ill-known-time-point}
\textbf{Ill-known time point.}\\
Consider a time domain $\T$; the uncertainty about the values of the ill-known time point $X$ is given by the possibility distribution $\pi_X$:
\begin{equation}
\label{eq:ill-known-time-point}
\Pi(X) = \pi_X(t) \in \left[0,1\right] , t \in \T
\end{equation}
\end{definition}

It is also possible to specify an ill-known time point by a convex combination of ill-known constraints, as shown by equation \eqref{eq:ill-known-value-def-by-const-unc}.

@@

\subsubsection{\label{subsubsec:ill-known-time-datatypes}Datatypes}
The data type for the representation of an ill-known time point allows the representation of the values shown in Table \ref{tbl:time-point-types}
%The datatypes for an  $X$ are shown in Table \ref{}.
\vglue13pt
%\begin{table}[htbp]
\tcap{\label{tbl:time-point-types}Values for the time point data type.}
\centerline{\small TIME POINT DATA TYPES}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c p{0.4\columnwidth} p{0.3\columnwidth}}\\
Subtype & Value & Representation \\
\hline
1 & A single time point &  $1/x, x \in \T$\\
2 & A possibility distribution in the numeric domain & A fuzzy number or a fuzzy interval.\\
3 & An unknown value & \textbf{UNKNOWN}$= \lbrace1/t,\forall t \in \T \rbrace$\\
4 & An undefined value & \textbf{UNDEFINED}$= \lbrace 0/t,\forall t \in \T \rbrace$\\
5 & A null value & \textbf{NULL} $=\lbrace 1/$Unknown, 1/Undefined $\rbrace$\\
\hline\\
\end{tabular}
}

%Example of different uses of the ill-known time value.
\begin{example}
Consider a historical database with data from medieval diplomatic documents. The following fields are stored: the digital identifier \emph{ID} which is the primary key  and the estimated time when the document was sent (field \emph{Date}). Table \ref{tbl:sample-time-point} contains some example data from this database. 
\end{example}

@@

%\begin{definition}
\textbf{Ill-known time interval}\\
An ill-known time interval denoted by $\left[X, Y\right]$ is a precise time interval whose boundaries are not precisely known. The ill-known time interval is defined by two ill-known  time points, namely $X$,$Y$. The evaluation of the ill-known time interval is given by equations \eqref{ill-known-pos},\eqref{ill-known-nec}.

% In order to know whether all the points in the crisp time interval $I=\left[a, b\right]$ are within the boundaries of the ill-known time interval $\left[X, Y\right]$ we define the following two ill-known constraints $C_{1}, C_{2}$ 
% \begin{eqnarray}
% \label{eq:constraint-c1}
% C_1\stackrel{\triangle}{=}\left(\geq,X\right)  \\
% \label{eq:constraint-c2}
% C_2\stackrel{\triangle}{=}\left(\leq,Y\right)
% \end{eqnarray}
% Note that the boolean operator applied to the constraints is $\bool=\wedge$. Hence, the construct of both constraints is $C_1 \wedge C_2$
% This means that we want to know if all the points in the interval $I$ are greater than or equal to $X$ and, at the same time, smaller than or equal to $Y$.
% %\end{definition}
% 
% %% equations for the poss and nec!!
% The simplified equations for the possibility and the necessity measures are~\cite{Pons2011}:
% \begin{eqnarray}
% \label{eq:time-interval-pos}
% \Pos\left(\lambda([a,b])\right)&=&\\
% \nonumber
% \min\bigg(\sup_{a\leq w}\pi_{X}(w),\sup_{b\geq w}\pi_{Y}(w)\bigg)\\
% \label{eq:time-interval-nec}
% \Nec\left(\lambda([a,b])\right)&=&\\
% \nonumber
% \min\bigg(\inf_{a>w}1-\pi_{X}(w),\inf_{b<w}1-\pi_{Y}(w)\bigg).
% \end{eqnarray}

@@

\subsubsection{\label{subsubsec:representation-semi-open}Representation of semi-open time intervals}
As mentioned before, the problem resides in the representation of this kind of interval.

% In order to get a short representation for these intervals, two constants are defined: $FB$ -\emph{From the Beginning}- it is only valid for the left value of the interval and $UC$ -\emph{Until Changed}- (this name is usual in the temporal database community, see ~\cite{Jensen1994}) which is only valid for the right value of the interval. Note that these two constants are aliases for a function that obtains both possibility and necessity measures.

Because of the ill-known constraints $C_{1}, C_{2}$, a function called \emph{Open} should be defined in order to deal with a proper representation of these intervals.

\begin{definition}
\label{def:open-func}
$Open(C)$\\
Consider an ill-known value $T$, a binary relationship $B_r$ one of $\left(\leq, <, >, \geq\right)$ and the constraint $C\stackrel{\triangle}{=}\left(R,T\right)$.
The function $\Open(C) = \left(I_p(C), I_n(C) \right)$ provides both possibility and necessity measures for all the points in the open part of a semi-open ill-known time interval.

The possibility and necessity measures are defined by:
\begin{eqnarray}
\label{eq:open-pos-nec}
I_p \left(C \right) &=& \Big(\sup_{r\in\T \Rp\  w}\pi_{T}(w)\Big)\\
I_n \left(C \right) &=& \Big(\inf_{r\in\T \Rn\  w}1-\pi_{T}(w)\Big)
\end{eqnarray}
Where the values for the binary relations $\Rp$ and $\Rn$ are shown in Table \ref{tbl:open-pos-nec-rels}.
\end{definition}

\vglue13pt
%\begin{table}[htbp]
\tcap{Relations for the $\Open(C)$ function. Depending on the relation $B_r = \left(\leq, <, >, \geq\right)$ in the constraint $C$, the values for $\Rp$ and $\Rn$ are shown.}
\centerline{\small Relations}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c c c}\\
\hline
Constraint & $\Rp$ & $\Rn$\\ \hline
$C\stackrel{\triangle}{=}\left(<,T\right)$ & $>$ & $\leq$ \\
$C\stackrel{\triangle}{=}\left(\leq,T\right)$ & $\geq$ & $<$ \\
$C\stackrel{\triangle}{=}\left(>,T\right)$ & $<$ & $\geq$ \\
$C\stackrel{\triangle}{=}\left(\geq,T\right)$ & $\leq$ & $>$ \\
\hline\\
\end{tabular}
\label{tbl:open-pos-nec-rels}
}

@@

\begin{example}
\label{ex:open-time-interval}
Consider an ill-known time interval given by $\left[\FB, Y \right]$. Consider also that in this case, $Y= \left[15/10/2012,3,4 \right] $. Figure \ref{fig:example-open-time-interval} shows the representation for $Y$. The user wants to obtain the possibility and the necessity measures for the $\FB$ part of the interval.
\end{example}

\begin{eqnarray}
\nonumber
\FB = \Open(C_2) \mbox{ with } C_2\stackrel{\triangle}{=}\left(\leq,Y\right)\\
\nonumber
I_p(C_2)= \Big(\sup_{r\in\T \geq\  w}\pi_{Y}(w)\Big)\\
\nonumber
I_n(C_2) = \Big(\inf_{r\in\T <\  w}1-\pi_{Y}(w)\Big)
\end{eqnarray}

@@

\subsubsection{\label{subsubsec:ill-known-time-interval-datatypes}Datatypes}
In order to properly represent  an ill-known time interval in a database, some datatypes are needed. Because of the ill-known constraints, not all the combinations of datatypes for each ill-known time point (see Table \ref{tbl:time-point-types}) are allowed. Table \ref{tbl:time-interval-types} shows all the possible datatypes combinations for the  for an ill-known time interval denoted by $\left[X, Y\right]$.

@@

% \subsection{\label{subsec:fuzzy-allen-relations}Allen's relations}
% The Allen's relations can be represented by a Boolean combination of ill-known constraints. The constructs for the constraints are shown in Table \ref{tbl:fuzzy-allen-relations}. The possibility and necessity measures are obtained for each Boolean function as explained in Section \ref{subsec:interval-evaluation-by-ill-known-constraints} in equations \eqref{eq:conjunctive1}-\eqref{eq:negation}.
% 
% 
% \vglue13pt
% %\begin{table}[htbp]
% \tcap{\label{tbl:fuzzy-allen-relations}Constructs of constraints related to their respective Allen relations, as used in the presented work. In this table, the ill-known time interval $J = \left[X, Y\right]$ in a record $r$ has a start point described by possibilistic variable $X$ and an end point described by possibilistic variable $Y$. The crisp time interval in the user's temporal demand is denoted $I$.}
% \centerline{\small Relations}
% \vglue-6pt
% \centerline{\small\baselineskip=13pt
% \begin{tabular}{p{0.2\columnwidth} p{0.8\columnwidth} }\\
% \hline
% Allen Relation & Construct of constraints\\
% \hline
% I before J & $C_1 \triangleq \left(<,X\right)$ \\
% \hline
% I equal J & $\left(C_1 \triangleq \left(\geq,X\right)\right) \wedge \neg \left(C_2 \triangleq \left(\neq,X\right) \right) \wedge \left(C_3 \triangleq \left(\leq,Y\right) \right) \wedge \neg \left(C_4 \triangleq \left(\neq,Y\right)\right)$ \\
% \hline
% I meets J & $\left(C_1 \triangleq \left(\leq,X\right)\right) \wedge \neg \left(C_2 \triangleq \left(\neq,X\right) \right)$ \\
% \hline
% I overlaps J & $\left(C_1 \triangleq \left(<,Y\right)\right) \wedge \neg \left(C_2 \triangleq \left(\leq,X\right) \right) \wedge \neg \left(C_3 \triangleq \left(\geq,X\right) \right)$ \\
% \hline
% I during J & $\left( \left (C_1 \triangleq \left(>,X\right) \right) \wedge \left(C_2 \triangleq \left(\leq,Y\right) \right) \right) \vee \left( \left(C_3 \triangleq \left(\geq,X\right) \right) \wedge  \left(C_4 \triangleq \left(<,Y\right) \right) \right)$ \\
% \hline
% I starts J & $\left(C_1 \triangleq \left(\geq,X\right) \right) \wedge \neg \left(C_2\triangleq \left(\neq,X\right)\right)$ \\
% \hline
% I finishes J & $\left(C_1 \triangleq \left(\leq,Y\right) \right) \wedge \neg \left(C_2\triangleq \left(\neq,Y\right)\right)$ \\
% \hline\\
% \end{tabular}
% }
